\papertitle

### 2017-03-31 article-icmc2017.tex
update

\usepackage{icmc2017template}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{soul}
\usepackage{microtype}
\usepackage{csquotes}
\usepackage{polyglossia}
\setmainfont{Times New Roman}

@@

For instance, given two processes $p_1, p_2$, what can be said of a program where $p_1$ executes from $t=0$ to $t=5$ seconds, and $p_2$ from $t=3$ to $t=6$ when $p_1$ and $p_2$ operate on the same data.

@@

The simplest strategy can be to only allow execution when all processes are active and have been given a definite order of execution: in this case the processes will only execute from $t=3$ to $t=5$ and behave like function composition. 
While safe from the point of view of software execution, we will show that other execution policies leveraging information from the temporal layout of the processes can provide composers with new creative capabilities, through an example that uses Pure Data.

@@

@@

\subsection{Temporal formalism}
This work implicitly uses the OSSIA formalism~\cite{celerier2015ossia} in its examples to provide primitives relative to the evolution of time: time constraints (horizontal lines), time nodes (vertical lines). 
The elements of its syntax are presented in fig.~\ref{fig.iscore-example}.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=0.35\textwidth]{images/iscore-example.eps}
  \caption{Part of an OSSIA scenario, showcasing the temporal syntax used. 
    A full horizontal line means that the time must not be interrupted, 
    while a dashed horizontal line means that the time of this time constraint can be interrupted to proceed 
  to the following parts of the score according to an external event.}
  \label{fig.iscore-example}
\end{figure}

\section{Definitions}
We call process an entity that consists in a function $f$, an associated state, any number of inputs and output ports, and an activation status.
$f$ has a single argument $t \in \mathbb{R^+}$.
The execution of $f$ may read any number of tokens from the inputs and write any number of tokens to the outputs.

@@

\begin{itemize}
  \item Synchronous dataflows between processes: input and output ports can be connected together to enforce data dependencies.
  \item Temporal graphs, based on the OSSIA formalism: $G_t(V_t, E_t)$ where $V_t$ is the set of time nodes and $E_t$ is the set of time constraints in a score. 
  For coherency, we call these graphs temporal graphs.
\end{itemize}

@@

There may be more than a single temporal graph: they are defined as processes themselves, which provides them with time information.
A temporal graph does not have inputs or outputs.

@@

\begin{figure}[h]
	\centering
	\begin{subfigure}{0.15\textwidth}\centering
		\begin{tikzpicture}[scale=0.5]
		\node[draw, circle] (f1) at (0, 0) { $f_1$ };
		\node[draw, circle] (f2) at (3, 0) { $f_2$ }
		edge [<-] (f1);
		\end{tikzpicture}\caption{}\label{fig.f1tof2}
	\end{subfigure}~
	\begin{subfigure}{0.15\textwidth}\centering	
		\begin{tikzpicture}[scale=0.5]
		\node[draw, circle] (f1) at (0, 0) { $f_1$ };
		\node[draw, circle] (f2) at (3, 0) { $f_2$ }
		edge [->] (f1);
		\end{tikzpicture}\caption{}
	\end{subfigure}~	
	\begin{subfigure}{0.15\textwidth}\centering
		\begin{tikzpicture}[scale=0.5]
		\node[draw, circle] (f1) at (0, 0) { $f_1$ };
		\node[draw, circle] (f2) at (3, 0) { $f_2$ };
		\end{tikzpicture}\caption{}
	\end{subfigure}
	\caption{Possible dataflows between $f_1$ and $f_2$. 
		As can be seen in fig.~\ref{fig.tgraph}, there are no relationships between $f_1$ and $f_2$ in the temporal graph.}
	\label{fig.simple-dataflow}
\end{figure}

@@

\centering
\begin{tikzpicture}[]
\input{scores/score.tex}
\end{tikzpicture}
\caption{The matching temporal graph}
\label{fig.tgraph}
\end{figure}

\begin{figure}
\centering
\begin{subfigure}{0.22\textwidth}
  \centering
  \begin{tikzpicture}[scale=0.25]
    \draw[] (0, 0) rectangle (5, 2) node[pos=.5] { $f_1$ };
    \draw[] (0 + 3, 0 - 2.5) rectangle (5 + 3, 2 - 2.5) node[pos=.5] { $f_2$ };

@@

% TODO sample-precision of the activation / deactivation of a process / node
% TODO 
\section{Relationships}
\label{sec.relationships}
We are interested in the relationships between nodes of the dataflow graph when they produce compatible tokens, whether the production is specified implicitly or explicitly, and when taking into account deactivated nodes.

@@

The delayed connection can behave in two ways:
\begin{itemize}
  \item Readers of the buffer always start from the same point: the beginning of the previous function in the callback chain. 
  The frame pointer would be located in each delayed connection.
  \item Readers of the buffer continue from the latest read position. 
  The frame pointer would be located in the source port.

@@

This relationship has to be specified explicitly to allow for the tokens to be bufferised.

@@

We define also define a strictness level for the delayed connection. 
In the strict case, a node will only be able to execute if the source has produced enough tokens.

@@

\node (t1) at (1.45, -3.45) {\footnotesize $t_0$};
\node (t2) at (4, -3.45) {\footnotesize $t_1$};
\node (t3) at (6.5, -3.45) {\footnotesize $t_2$};
\end{tikzpicture}
\caption{Another execution trace. $f_2$'s input corresponds to $f_1$'s output, delayed: this creates a causality problem.}
\label{fig.delay}
\end{figure}

@@

\section{Default behaviors for orders}
\label{sec.order}
We assess here the various mechanisms possible to specify the dependencies between nodes of the dataflow.
Information provided provided in the temporal graph can be leveraged to connect the nodes at runtime in a fashion 
that would make sense to composers \& authors.

@@

Given a global environment with the variables: 
\begin{lstlisting}
/foo/bar/ba     1.0
/foo.1/bob/bu   2.0
/foo.2/bar/be  -1.0
/foo.42/bar/bo -3.0
\end{lstlisting}

Such a pattern would yield the tokens $-1.0$ and $-3.0$ at the input of the port matching the pattern.

@@

\section{Applied example: generative music}
\begin{figure}[h]
    \centering
    \begin{subfigure}{0.21\textwidth}\centering
        \includegraphics[scale=0.5]{images/pd-1.png}
        \caption{$f_1, f'_1$: generates a random note on each bang}
    \end{subfigure}
    \begin{subfigure}{0.21\textwidth}\centering
        \includegraphics[scale=0.5]{images/pd-2.png}
        \caption{$f_2$: generates another random note on each bang}
    \end{subfigure}

    \begin{subfigure}{0.21\textwidth}\centering
        \includegraphics[scale=0.5]{images/pd-3.png}
        \caption{$f_3$: synthesizes a note, sets the volume}
    \end{subfigure}
    \begin{subfigure}{0.21\textwidth}\centering
        \includegraphics[scale=0.5]{images/pd-4.png}
        \caption{$f_4$: applies an high-pass filter}
    \end{subfigure}
    \begin{subfigure}{0.21\textwidth}\centering
        \includegraphics[scale=0.5]{images/pd-5.png}
        \caption{$f_5, f'_5$: applies a low-pass filter}
    \end{subfigure}
    \caption{The functions we use in our example, as defined in PureData.}
    \label{fig.pdpatches}
\end{figure}

@@

    \caption{The data relationships between instances of $f_{1,..5}$ in the program.}
    \label{fig.datapatch}
\end{figure}

@@

    The sound that can be heard is a single tone with some amount of high-pass filtering.
    \item $f_2$ starts: random notes will be generated and put at the \lstinline|/note| input of $f_3$ directly, bypassing the environment.

    \item $f_5$ starts: the low-pass filter adds itself after the high-pass filter.
    \item $f_2$ stops: the sound reverts to a continuous tone (the last that was written, again).
    \item $f'_1$ starts: tones are generated again, but with a faster rhythm.
    \item $f_4$ stops: there is no sound anymore due to the strict dependency. $f_5$ applies its low-pass filter to silence.
    \item $f_5$ stops, and just afterwards $f'5$ starts. Since $f'5$ was in a delayed relationship with $f_4$, the sound produced corresponds to the sound that was heard at the beginning of $f_4$, with both $f_4$'s high pass filtering, and $f'5$'s low-pass filtering.
\end{enumerate}
\section{Conclusion}
Throughout this paper, the effects of adding macro-temporal semantics to dataflows were studied. 
The first step is to introduce an environment that allows execution of a program to continue if not all the nodes of a dataflow are currently being executed.
Then, we study the new possibilities offered by the presence of the environment by the different ways to interpret the absence of execution of a node.
This includes matching nodes together through pattern matching, and distinct policies of execution that can take into account the evolution of the program in time.

>>>>

\usepackage{icmc2017template}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{soul}
\usepackage{microtype}
\usepackage{csquotes}
\usepackage{polyglossia}
\usepackage{footmisc}
\setmainfont{Times New Roman}

@@

For instance, given two processes $p_1, p_2$, what can be said of a program where $p_1$ executes from $t=0$ to $t=5$ seconds, and $p_2$ from $t=3$ to $t=6$ when $p_1$ and $p_2$ operate on the same data?

@@

The simplest strategy can be to only allow execution when all processes are active and have been given a definite order of execution: in this case the overall program will only execute during the intersection of the activation of $p_1$ and $p_2$, which goes from $t=3$ to $t=5$. 
While safe from the point of view of software execution, we will show that other execution policies leveraging information from the temporal layout of the processes can provide composers with new creative capabilities, through an example that uses Pure Data.

@@

\begin{figure} [h]
	\centering
	\includegraphics[width=0.35\textwidth]{images/iscore-example.eps}
	\caption{Part of an OSSIA scenario, showcasing the temporal syntax used. 
		A full horizontal line means that the time must not be interrupted, 
		while a dashed horizontal line means that the time of this time constraint can be interrupted to proceed 
		to the following parts of the score according to an external event.}
	\label{fig.iscore-example}
\end{figure}

@@

\subsection{Temporal formalism}

This work implicitly uses the OSSIA formalism~\cite{celerier2015ossia} in its examples to provide primitives relative to the evolution of time: time constraints (horizontal lines), time nodes (vertical lines). 
The elements of its syntax are presented in fig.~\ref{fig.iscore-example}.

@@

\section{Definitions}
\begin{figure}[h]
	\centering
	\begin{subfigure}{0.15\textwidth}\centering
		\begin{tikzpicture}[scale=0.5]
		\node[draw, circle] (f1) at (0, 0) { $f_1$ };
		\node[draw, circle] (f2) at (3, 0) { $f_2$ }
		edge [<-] (f1);
		\end{tikzpicture}\caption{}\label{fig.f1tof2}
	\end{subfigure}~
	\begin{subfigure}{0.15\textwidth}\centering	
		\begin{tikzpicture}[scale=0.5]
		\node[draw, circle] (f1) at (0, 0) { $f_1$ };
		\node[draw, circle] (f2) at (3, 0) { $f_2$ }
		edge [->] (f1);
		\end{tikzpicture}\caption{}
	\end{subfigure}~	
	\begin{subfigure}{0.15\textwidth}\centering
		\begin{tikzpicture}[scale=0.5]
		\node[draw, circle] (f1) at (0, 0) { $f_1$ };
		\node[draw, circle] (f2) at (3, 0) { $f_2$ };
		\end{tikzpicture}\caption{}
	\end{subfigure}
	\caption{Possible dataflows between $f_1$ and $f_2$. 
		As can be seen in fig.~\ref{fig.tgraph}, there are no relationships between $f_1$ and $f_2$ in the temporal graph.}
	\label{fig.simple-dataflow}
\end{figure}

We call process an entity that consists in a function $f$, an associated state, any number of inputs and output ports, and an activation status.
$f$ has a single argument $t \in \mathbb{R^+}$.
The execution of $f$ may read any number of tokens from the inputs and write any number of tokens to the outputs.

@@

\begin{itemize}
  \item Synchronous dataflow graphs between processes:~\\$G_d(V_d, E_d)$ where $V_d$ is the set of processes and $E_d$ the set of data bindings between processes.

  \item Temporal graphs, based on the OSSIA formalism: $G_t(V_t, E_t)$ where $V_t$ is the set of time nodes and $E_t$ is the set of time constraints in a score. 
  For coherency, we call these graphs temporal graphs.
\end{itemize}

@@

A program may consist in more than a single temporal graph: they are defined as processes themselves, which provides them with time information.

@@

@@

\centering
\begin{tikzpicture}[]
\input{scores/score.tex}
\end{tikzpicture}
\caption{The example temporal graph we use: both $f_1$ and $f_2$ can start at any time. $f_1$ and $f_2$ both have a given, finite duration.}
\label{fig.tgraph}
\end{figure}

\begin{figure}[htp]
\centering
\begin{subfigure}{0.22\textwidth}
  \centering
  \begin{tikzpicture}[scale=0.25]
    \draw[] (0, 0) rectangle (5, 2) node[pos=.5] { $f_1$ };
    \draw[] (0 + 3, 0 - 2.5) rectangle (5 + 3, 2 - 2.5) node[pos=.5] { $f_2$ };

@@

% TODO sample-precision of the activation / deactivation of a process / node
% TODO 
\section{Relationships}
\label{sec.relationships}
We are interested in the relationships between nodes of the dataflow graph when they produce compatible tokens, whether the production is specified implicitly or explicitly, and when taking into account deactivated nodes.
That is, given a node of the dataflow executing and producing tokens, we must define which following nodes, if any, will receive the tokens and when will they execute.

@@

The delayed connection can behave in two ways:
\begin{itemize}
  \item Readers of the buffer always start from the same point: the beginning of the previous function in the callback chain. 
  The frame pointer would be located in each delayed connection, and would not be shared between processes.
  \item Readers of the buffer continue from the latest read position. 
  The frame pointer would be located in the source port, and would be shared across all processes reading from it.

@@

This relationship has to be specified explicitly to allow for the tokens to be buffered.

@@

We also define a strictness level for the delayed connection. 
In the strict case, a node will only be able to execute if the source has produced enough tokens.

@@

\node (t1) at (1.45, -3.45) {\footnotesize $t_0$};
\node (t2) at (4, -3.45) {\footnotesize $t_1$};
\node (t3) at (6.5, -3.45) {\footnotesize $t_2$};
\end{tikzpicture}
\caption{Another execution trace, stemming from a different score than the one presented in fig.~\ref{fig.tgraph}. $f_2$'s input corresponds to $f_1$'s output, delayed: this creates a causality problem. }
\label{fig.delay}
\end{figure}

@@

\section{Default behaviors for orders}
\label{sec.order}
We assess here the various mechanisms possible to specify the dependencies between nodes of the dataflow.
Information provided in the temporal graph can be leveraged to connect the nodes at runtime in a fashion 
that would make sense to composers \& authors.

@@

Given a global environment with the variables:~\\
\begin{tabular}{c}
\begin{lstlisting}
/foo/bar/ba     1.0
/foo.1/bob/bu   2.0
/foo.2/bar/be  -1.0
/foo.42/bar/bo -3.0
\end{lstlisting}
\end{tabular}~\\~\\
Such a pattern would yield the tokens $-1.0$ and $-3.0$ at the input of the port matching the pattern.

@@

\section{Applied example: generative music}
\begin{figure}[h]
    \centering
    \begin{subfigure}{0.21\textwidth}\centering
        \includegraphics[scale=0.5]{images/pd-1.png}
        \caption{$f_1, f'_1$: note generator.}
    \end{subfigure}
    \begin{subfigure}{0.21\textwidth}\centering
        \includegraphics[scale=0.5]{images/pd-2.png}
        \caption{$f_2$: note generator, at a lower tempo.}
    \end{subfigure}

    \begin{subfigure}{0.21\textwidth}\centering
        \includegraphics[scale=0.5]{images/pd-3.png}
        \caption{$f_3$: synthesizes the input note, sets the volume.}
    \end{subfigure}

    \begin{subfigure}{0.21\textwidth}\centering
        \includegraphics[scale=0.5]{images/pd-4.png}
        \caption{$f_4$: high-pass filter.}
    \end{subfigure}
    \begin{subfigure}{0.21\textwidth}\centering
        \includegraphics[scale=0.5]{images/pd-5.png}
        \caption{$f_5, f'_5$: low-pass filter.}
    \end{subfigure}
    \caption{The functions we use in our example, as defined in Pure Data.}
    \label{fig.pdpatches}
\end{figure}

@@

    \caption{The dataflow relationships between instances of $f_{1,..5}$ in the program. The global environment contains variables for \lstinline|/vol| and \lstinline|/filt| when the execution starts.}
    \label{fig.datapatch}
\end{figure}

@@

    The sound that can be heard is a single tone with some amount of high-pass filtering.
    \item $f_2$ starts: random notes will be generated at a lower tempo and put at the \lstinline|/note| input of $f_3$ directly, bypassing the environment.

    \item $f_5$ starts: the low-pass filter adds itself after the high-pass filter.
    \item $f_2$ stops: the sound reverts to a continuous tone (the last that was written, again).
    \item $f'_1$ starts: tones are generated again, but with a faster tempo.
    \item $f_4$ stops: there is no sound anymore due to the strict dependency. $f_5$ applies its low-pass filter to silence.
    \item $f_5$ stops, and just afterwards $f'5$ starts. Since $f'5$ was in a delayed relationship with $f_4$, the sound produced corresponds to the sound that was heard at the beginning of $f_4$, with both $f_4$'s high pass filtering, and $f'5$'s low-pass filtering.
\end{enumerate}
\section{Conclusion}
Throughout this paper, the effects of adding macro-temporal semantics to dataflows were studied. 
The first step is to introduce an environment that allows execution of a program to continue if not all the nodes of a dataflow are currently being executed.
Then, we study the new possibilities offered by the presence of the environment by the different ways to interpret the absence of execution of a node.
This includes matching nodes together through pattern matching, and distinct policies of execution that can take into account the evolution of the program in time.
